Efficient pedigree recording for fast population genetics simulation

### 2017-11-19 forwards_paper.tex
Revise tables subhed per plr suggestion
Remove F, clean up a bit

%%%%%%
\subsection*{Recording a tree sequence}

@@

We begin in~\algref{W1} by allocating our initial population $P$
(a vector of $N$ node IDs),
and creating $N$ nodes with birth time $T$ generations ago, recorded in the
vector $\tau$. The set $E$ is used to store the edges that we output during the
simulation, and $n$ is the number of nodes created so far (and so, the ID of
the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
an individual $P_j$ has died in this generation. If it has, we replace it in
steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
individual. When an individual in the population dies, we first allocate
a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $n$
from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
$n$, ready to represent the next new node.

@@

At any point in a simulation scheme like Algorithm~\algref{W}, genealogical history is recorded
in a tree sequence. This has two additional advantages. First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime. 
This process of forward simulation coupled with sequential simplification is our
method

\jda{I didn't write down the full thing, but mightn't things be clearer if we
	have a statement of the combined method?}
\begin{taocpalg}{F}{Forwards-time simplifed tree sequence}{
Simulates a population starting with $N$ diploid individuals on a chromosome of length $L$ for $T$
generations.
}

\algstep{F0.}{Initialize nodes.}{
Generate initial haploid nodes $\mathcal N$:
1.) using coalsecent simulation with \msprime{} for $2N$ initial haploid samples $S$ on
a chromosome of length $L$ and taking $\mathcal N = \{\texttt{id}(u) : u \in S \}$; or
2.) simply defining $\mathcal N = \{1\dots 2N\}$ as the node ids.
Define a mapping from  $\mathcal M: \{1\dots 2N\} \to \mathcal N$. In the second
case $\mathcal M$ is the identity.
}

\algstep{F1.}{Initialize edges.}{
Define the initial edges $E_0$: 1) either from the coalescent simulation's final
state, or 2) as \dots
}

\algstep{F2.}{Record trees till time $T - \Delta$.}{ 
Simulate, using Algorithm~\algref{W} or another algorithm, $2N$ haploid
individuals using $\mathcal M$ to renumber the nodes, producing nodes $n$ and
edgesets $E$.
}

\algstep{F3.}{Simplification over $\Delta$.}{
Select final samples $S'$ and using the simplification Algorithm~\algref{S},
simplify $n$ and $E$ to produce a simplified tree sequence.
}

\algstep{F4.}{Continue?}{
	Stop if $T - \delta  = 0$, else continue.
}
\end{taocpalg}

@@

\jda{I'd like to outline what we're doing here with reference to alg F}
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
using Algorithm~\algref{W}
for $T$ generations,
with the probability of death, $\delta$, set to 1 for simplicity.
Since there is exactly one crossing over every generation,
this produces tables of 
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that 
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and three times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 4,700.

>>>>

%%%%%%
\subsection*{Working with a tree sequence through tables}

@@

\jda{TODO: rewrite with tables}
We begin in~\algref{W1} by allocating our initial population $P$
(a vector of $N$ node IDs),
and creating $N$ nodes with birth time $T$ generations ago, recorded in the
vector $\tau$. The set $E$ is used to store the edges that we output during the
simulation, and $n$ is the number of nodes created so far (and so, the ID of
the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
an individual $P_j$ has died in this generation. If it has, we replace it in
steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
individual. When an individual in the population dies, we first allocate
a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $n$
from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
$n$, ready to represent the next new node.

@@

At any point in a simulation scheme that, like Algorithm~\algref{W}, records
data into tables the genealogical history is recorded
in a tree sequence.
This has two additional advantages, as seen in Algorithm~\algref{W}.
First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime. 
We provide results for simulation timing from two specific implementations of this
method (forward simulation coupled with sequential simplification)
in the next section, but first we analyze the expected resource use of the
general Algorithm~\algref{W}.

@@

\jda{It'd be good to outline what readers should expect here.}
\jda{and note we're doing the space analysis for with reference to alg W with $s=1$}
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
using Algorithm~\algref{W}
for $T$ generations,
with the probability of death, $\delta$, set to 1 for simplicity.
Since there is exactly one crossing over every generation,
this produces tables of 
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that 
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and three times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 4,700.
